---
bibliography: references.bib
---

# An introduction to the United States Census and the R programming language

```{r setup-ch1, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Census data: an overview

Basically, describe decennial and ACS data in general terms, and give a nod to some of the other surveys/datasets.

## Census hierarchies

Explain the Census geographic hierarchy and how data are tabulated

## Focus of the book

## What is R?

A nod to the history of R and contemporary ecosystems (e.g. tidyverse, R spatial).

## R and the Census API

Show how to make a simple request to the Census API with httr and jsonlite. Discuss the concept of an API key.

## R packages for Census data analysis

A robust ecosystem has emerged in the R community for acquiring, analyzing, and working with data from the US Census Bureau. R Census data packages generally make requests to the Census website for the user then format the data in a way that is suitable for data analysis. An overview of this ecosystem is found in Ari Lamstein and Logan Powell's [*A Guide to Working with US Census Data in R*](https://rconsortium.github.io/censusguide/). While all of the packages mentioned are of considerable merit, this book focuses on applications using a handful of those packages. **tidycensus** [@walker_and_herman2021] gives R users access to demographic and spatial data from the US Census Bureau in tidyverse-style format; this is the package most prominently featured in this book, and figures prominently in several chapters including 2, 3, 4, 6, and 9. **tigris** [@walker2021] is a direct interface to the US Census Bureau's geographic datasets including the TIGER/Line and cartographic boundary shapefiles; it is the focus of Chapter 5. The **ipumsr** package [@ellis_and_burk2020] helps R users work with data from the [IPUMS project](https://ipums.org/), which includes a wide range of harmonized Census datasets including recent and historical microdata, tabulated data, and geographic data. ipumsr is a focus of Chapter 11 of this book.

While these packages cover a wide range of use cases, several other R packages include additional functionality that will be very useful for the R Census data analyst. Some packages of note are covered in the following subsections.

### censusapi

```{r censusapi, eval = FALSE}
library(censusapi)

key <- "84dba8250c52c9a11ec14575dce9083eea98d7e9"

econ17 <- getCensus(
    name = "encbasic",
    vintage = 2017,
    vars = c("EMP", "OPTAX", "GEOTYPE"),
    region = "state:48",
    regionin = "county:113",
    naics2017 = 72,
    key = key)

head(ewks_2012)
```

### lehdr

```{r lehdr, eval = FALSE}
library(lehdr)


```

### choroplethr

```{r choroplethr}

```

### totalcensus

```{r totalcensus}

```
