# Spatial analysis with US Census data

A very common use-case of spatial data from the US Census Bureau is _spatial analysis_.  Spatial analysis refers to the performance of analytic tasks that explicitly incorporate the spatial properties of a dataset.  Principles in spatial analysis are closely related to the field of _geographic information science_, which incorporates both theoretical perspectives and methodological insights with regards to the use of geographic data.  

Traditionally, geographic information science has been performed in a _geographic information system_, or “GIS”, which refers to an integrated software platform for the management, processing, analysis, and visualization of geographic data.  As evidenced in this book already, R packages exist for handling these tasks, allowing R to function as a capable substitute for desktop GIS software like ArcGIS or QGIS.  

Traditionally, spatial analytic tasks in R have been handled by the __sp__ package and allied packages such as __rgeos__.  In recent years, however, the __sf__ package has emerged as the next-generation alternative to sp for spatial data analysis in R.  In addition to the simpler representation of vector spatial data in R, as discussed in previous chapters, sf also includes significant functionality for spatial data analysis that integrates seamlessly with tidyverse tools.  

This chapter covers how to perform common spatial analysis tasks with Census data using the sf package.  As with previous chapters, the examples will focus on data acquired with the tidycensus and tigris packages, and will cover common workflows of use to practitioners who work with US Census Bureau data.  

## Spatial overlay and spatial joins

Spatial data analysis allows practitioners to consider how geographic datasets interrelate in geographic space.  This analytic functionality facilitates answering a wide range of research and analytic questions that would otherwise prove difficult without reference to a dataset’s geographic properties.  

One common use-case employed by the geospatial analyst is _spatial overlay_.  Key to the concept of spatial overlay is the representation of geographic datasets as _layers_ in a GIS.  This representation is exemplified by the graphic below.  

{layers graphic}

In this representation, different components of the landscape that interact in the real world 

## Distance and proximity analysis



## Group-wise spatial data analysis


## Better cartography with spatial overlay

One of the major benefits of working with the [tidycensus](https://walkerke.github.io/tidycensus/) package to get Census data in R is its ability to retrieve pre-joined feature geometry for common geographies (`"state"`, `"county"`, `"tract"`, `"block group"`, `"block"`, and `"zcta"`) with the argument `geometry = TRUE`.  tidycensus uses the [tigris](https://github.com/walkerke/tigris) package to fetch these geometries, which default to the Census Bureau's [cartographic boundary shapefiles](https://www.census.gov/geo/maps-data/data/tiger-cart-boundary.html).  Cartographic boundary shapefiles are preferred to the core [TIGER/Line shapefiles](https://www.census.gov/geo/maps-data/data/tiger-line.html) in tidycensus as their smaller size speeds up processing and because they are pre-clipped to the US coastline.

However, there may be circumstances in which your mapping requires more detail.  A good example of this would be maps of New York City, in which even the cartographic boundary shapefiles include water area.  For example, take this example of median household income by Census tract in Manhattan (New York County), NY: 

```{r}
library(tidycensus)
library(mapview)
options(tigris_use_cache = TRUE)

ny <- get_acs(geography = "tract", 
              variables = "B19013_001", 
              state = "NY", 
              county = "New York", 
              geometry = TRUE)

# mapview(ny, zcol = "estimate", legend = TRUE)

```

As illustrated in the graphic, the boundaries of Manhattan include water boundaries - stretching into the Hudson and East Rivers.  In turn, a more accurate representation of Manhattan's land area might be desired.  To accomplish this, a tidycensus user can use the core TIGER/Line shapefiles instead, then erase water area from Manhattan's geometry.  

tidycensus allows users to get TIGER/Line instead of cartographic boundary shapefiles with the keyword argument `cb = FALSE`.  This argument will be familiar to users of the tigris package, as it is used by tigris to distinguish between cartographic boundary and TIGER/Line shapefiles in the package.  

```{r}
ny2 <- get_acs(geography = "tract", 
              variables = "B19013_001", 
              state = "NY", 
              county = "New York", 
              geometry = TRUE, 
              cb = FALSE)
```

Next, tools in the tigris and [sf](https://github.com/r-spatial/sf) package can be used to remove the water area from Manhattan's Census tracts.  sf allows users to "erase" one geometry from another, [akin to tools available in desktop GIS software](http://pro.arcgis.com/en/pro-app/tool-reference/analysis/erase.htm).  The `st_erase()` function defined below is not exported by the package, but is defined in the documentation for `st_difference()`.  

The geometry used to "erase" water area from the tract polygons is obtained by the `area_water()` function in tigris, making sure to choose the option `class = "sf"`.  

```{r}
library(sf)
library(tigris)

st_erase <- function(x, y) {
  st_difference(x, st_union(st_combine(y)))
}

ny_water <- area_water("NY", "New York", class = "sf")

ny_erase <- st_erase(ny2, ny_water)

```



## Spatial neighborhoods and weights matrices

The spatial capabilities of tidycensus also allow for exploratory spatial data analysis (ESDA) within R.  


```{r}
library(tidycensus)
library(tidyverse)
library(sf)
library(spdep)
options(tigris_use_cache = TRUE)

la <- get_acs(geography = "tract", state = "CA", 
              county = "Los Angeles", variables = c(medincome = "B19013_001"), 
              geometry = TRUE) %>%
  st_transform(26910)

head(la)

```

* Compute weights matrix

```{r}
lasp <- as(la, "Spatial")

nb <- poly2nb(lasp, queen = TRUE)

weights <- nb2listw(nb)

```



## Global and local spatial autocorrelation



